Clustering analysis and statistics

### 2019-04-24 notebooks/Clustering analysis.Rmd
minor changes

```{r libraries, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggtree)
library(cluster)
library(Biostrings)
library(ape)
library(tibble)
library(cowplot)
library(knitr)
library(png)
```

@@

```{r import data}
# Vector of excluded samples
ex_samples <- c(
  "2016-17-565-1-S",
  "2016-02-428-2-S",
  "2016-02-486-2-S",
  "2016-02-732-2-S",
  "2014-01-1741-1-S"
  )

@@

# cgMLST data
cgmlst_stats <- read.table("../data/chewbbaca/complete_results/mdata_stats.tsv",
                         sep = "\t",
                         header = TRUE,
                         stringsAsFactors = FALSE)

detailed_cgmlst_stats <- read.table("../data/chewbbaca/results_statistics.tsv",
                                    sep = "\t",
                                    header = TRUE,
                                    stringsAsFactors = FALSE)

@@

@@

metadata_species <- read.table("../data/chewbbaca/complete_results/metadata_species_specific_trees.txt",
                               sep = "\t",
                               header = TRUE)

palette3 <- c("0" = "grey95",
              # gyrA
              "S83L" = "#c6dbef",
              "S83A" = "#9ecae1",
              "D87N" = "#6baed6",
              "D87H" = "#4292c6",
              "D87Y" = "#2171b5",
              "D87G" = "#08519c",
              "S83L, D87N" = "#08306b",
              # parC
              "S80I" = "#a1d99b",
              "S80R" = "#74c476",
              "S57T" = "#41ab5d",
              "A56T, S80I" = "#238b45",
              "S80I, E84V" = "#005a32",
              # parE
              "D475E" = "#fdd0a2",
              "D463N" = "#fdae6b",
              "S458A" = "#fd8d3c",
              "L416F" = "#f16913",
              "H516Y" = "#d94801",
              "L488M, A512T" = "#8c2d04",
              # PMQR
              "qnrA1" = "#762a83",
              "qnrB19" = "#ef8a62",
              "qnrS1" = "#a6dba0",
              "qnrS2" = "#5aae61",
              "qnrS4" = "#1b7837",
              "qepA4" = "#80cdc1",
              # num
              "1" = "#440154FF",
              "2" = "#3B528BFF",
              "3" = "#21908CFF",
              "4" = "#5DC863FF",
              "5" = "#FDE725FF")

# Broiler tree
broiler_cgMLST <- read.table("../data/chewbbaca/complete_results/broiler_cgmlst.txt",
                   sep = "\t",
                   header = TRUE,
                   colClasses = "factor")

broiler_tree <- as.phylo(hclust(daisy(broiler_cgMLST, metric = "gower"), method = "average"))

broiler_tree_annotated <- ggtree(broiler_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#4575b4",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.01,
               size = 6)

broiler_tree_opened <- rotate_tree(open_tree(broiler_tree_annotated, 10), 92.5)

broiler_complete <- gheatmap(broiler_tree_opened,
         metadata_species,
         offset = 0.1,
         width = 0.5,
         colnames_offset_y = 0.75,
         colnames_position = "top",
         font.size = 5) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Pig tree
pig_cgMLST <- read.table("../data/chewbbaca/complete_results/pig_cgmlst.txt",
                   sep = "\t",
                   header = TRUE,
                   colClasses = "factor")

pig_tree <- as.phylo(hclust(daisy(pig_cgMLST, metric = "gower"), method = "average"))

pig_tree_annotated <- ggtree(pig_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#74add1",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.01,
               size = 5.5)

pig_tree_opened <- rotate_tree(open_tree(pig_tree_annotated, 10), 92.5)

pig_complete <- gheatmap(pig_tree_opened,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames_offset_y = 0.53,
         colnames_position = "top",
         font.size = 5) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Red fox tree
fox_cgMLST <- read.table("../data/chewbbaca/complete_results/fox_cgmlst.txt",
                   sep = "\t",
                   header = TRUE,
                   colClasses = "factor")

fox_tree <- as.phylo(hclust(daisy(fox_cgMLST, metric = "gower"), method = "average"))

fox_tree_annotated <- ggtree(fox_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#f46d43",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.01,
               size = 5.5)

fox_tree_opened <- rotate_tree(open_tree(fox_tree_annotated, 10), 92)

fox_complete <- gheatmap(fox_tree_opened,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames_offset_y = 0.17,
         colnames_position = "top",
         font.size = 5) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Wild bird tree
bird_cgMLST <- read.table("../data/chewbbaca/complete_results/bird_cgmlst.txt",
                   sep = "\t",
                   header = TRUE,
                   colClasses = "factor")

bird_tree <- as.phylo(hclust(daisy(bird_cgMLST, metric = "gower"), method = "average"))

bird_tree_annotated <- ggtree(bird_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#fdae61",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.01,
               size = 5.5)

bird_tree_opened <- rotate_tree(open_tree(bird_tree_annotated, 10), 92.5)

bird_complete <- gheatmap(bird_tree_opened,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames_offset_y = 0.38,
         colnames_position = "top",
         font.size = 5) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Total plot
tot_plot <- plot_grid(broiler_complete,
          pig_complete,
          fox_complete,
          bird_complete,
          nrow = 2,
          ncol = 2,
          align = "h")

include_graphics("../figures/species_dendrogram.png")
```

@@

@@

ggplot(complete_data, aes(Percent)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.2, color = "black",
                fill = "grey80") +
  stat_function(fun = dnorm,
                args = with(complete_data, c(mean = mean(Percent), sd = sd(Percent))),
                color = "red") +
  geom_segment(aes(x = complete_data$Percent[1],
                   xend = complete_data$Percent[1], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#4575b4") +
  geom_segment(aes(x = complete_data$Percent[2],
                   xend = complete_data$Percent[2], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#74add1") +
  geom_segment(aes(x = complete_data$Percent[3],
                   xend = complete_data$Percent[3], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#f46d43") +
  geom_segment(aes(x = complete_data$Percent[4],
                   xend = complete_data$Percent[4], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#fdae61") +
  geom_text(label = "p < 0.001",
            x = 83.30,
            y = 0.16,
            angle = 90) +
  labs(x = "Percent (%) distances < 0.1 or > 0.8",
       y = "Density") +
  ggtitle(label = "Distribution of percentages",
          subtitle = "Number of iterations: 1000") +
  theme(plot.title = element_text(hjust = 0))

@@

# References
[1] Alikhan N-F, Zhou Z, Sergeant MJ, Achtman M. A genomic overview of the population structure of Salmonella. Casadesús J, editor. PLOS Genet [Internet]. 2018 Apr 5;14(4):e1007261. Available from: https://dx.plos.org/10.1371/journal.pgen.1007261

[2] Machado MP, Silva M, Carriço JA, Silva DN, Santos S, Ramirez M, et al. chewBBACA: A complete suite for gene-by-gene schema creation and strain identification. Microb Genomics. 2018;1–7.

>>>>

```{r libraries, include = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggtree)
library(cluster)
library(Biostrings)
library(ape)
library(tibble)
library(cowplot)
library(knitr)
library(distanceR)
```

@@

```{r import data}
# Vector of excluded samples
ex_samples <- c(
  "2016-17-565-1-S",
  "2016-02-415-2-S",
  "2016-02-428-2-S",
  "2016-02-486-2-S",
  "2016-02-732-2-S",
  "2014-01-1741-1-S"
  )

@@

# cgMLST data
cgmlst_stats <- read.table("../data/chewbbaca/complete_results/mdata_stats.tsv",
                         sep = "\t",
                         header = TRUE,
                         stringsAsFactors = FALSE) %>%
  filter(FILE != "2016-02-415_S208_pilon_spades.fasta")

detailed_cgmlst_stats <- read.table("../data/chewbbaca/results_statistics.tsv",
                                    sep = "\t",
                                    header = TRUE,
                                    stringsAsFactors = FALSE) %>%
  filter(Genome != "2016-02-415_S208_pilon_spades.fasta")

@@

names(tree_heatmap_data) <- c("GyrA","ParC","ParE","RpoB","MarA","MarR","SoxR","qepA4","qnr")

@@

metadata_species <- read.table("../data/chewbbaca/complete_results/metadata_species_specific_trees.txt",
                               sep = "\t",
                               header = TRUE) %>%
  select(gyrA, parC, parE, PMQR) %>%
  dplyr::rename("GyrA" = gyrA,
         "ParC" = parC,
         "ParE" = parE)

palette3 <- c("0" = "grey95",
              # gyrA
              "S83L" = "#c6dbef",
              "S83A" = "#9ecae1",
              "D87N" = "#6baed6",
              "D87H" = "#4292c6",
              "D87Y" = "#2171b5",
              "D87G" = "#08519c",
              "S83L, D87N" = "#08306b",
              # parC
              "S80I" = "#a1d99b",
              "S80R" = "#74c476",
              "S57T" = "#41ab5d",
              "A56T, S80I" = "#238b45",
              "S80I, E84V" = "#005a32",
              # parE
              "D475E" = "#fdd0a2",
              "D463N" = "#fdae6b",
              "S458A" = "#fd8d3c",
              "L416F" = "#f16913",
              "H516Y" = "#d94801",
              "L488M, A512T" = "#8c2d04",
              # PMQR
              "qnrA1" = "#dadaeb",
              "qnrB19" = "#bcbddc",
              "qnrS1" = "#9e9ac8",
              "qnrS2" = "#807dba",
              "qnrS4" = "#6a51a3",
              "qepA4" = "#4a1486")

# Broiler tree

broiler_tree <- calc_tree("../data/chewbbaca/complete_results/cgMLST_broiler.tsv")

broiler_tree_annotated <- ggtree(broiler_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#4575b4",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.02,
               size = 6)

broiler_complete <- gheatmap(broiler_tree_annotated,
         metadata_species,
         offset = 0.1,
         width = 0.5,
         colnames = FALSE) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Pig tree
pig_tree <- calc_tree("../data/chewbbaca/complete_results/cgMLST_pig.tsv")

pig_tree_annotated <- ggtree(pig_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#74add1",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.02,
               size = 5.5)

pig_complete <- gheatmap(pig_tree_annotated,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames = FALSE) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Red fox tree
fox_tree <- calc_tree("../data/chewbbaca/complete_results/cgMLST_fox.tsv")

fox_tree_annotated <- ggtree(fox_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#f46d43",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.02,
               size = 5.5)

fox_complete <- gheatmap(fox_tree_annotated,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames = FALSE) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Wild bird tree
bird_tree <- calc_tree("../data/chewbbaca/complete_results/cgMLST_bird.tsv")

bird_tree_annotated <- ggtree(bird_tree,
                                 layout = "circular",
                                 color = "#808080",
                                 size = 0.5) %<+% tree_metadata +
  geom_tippoint(color = "#fdae61",
                size = 4) +
  geom_tiplab2(aes(label = ST),
               offset = 0.02,
               size = 5.5)

bird_complete <- gheatmap(bird_tree_annotated,
         metadata_species,
         offset = 0.09,
         width = 0.5,
         colnames = FALSE) +
  scale_fill_manual(values = palette3) +
  guides(fill = FALSE)

# Total plot
tot_plot <- plot_grid(broiler_complete,
          pig_complete,
          fox_complete,
          bird_complete,
          nrow = 2,
          ncol = 2,
          align = "h",
          labels = c("Broiler","Pig","Red fox","Wild bird"))

include_graphics("../figures/species_dendrogram2.png")
```

# Minimum spanning tree
A minimum spanning tree calculated from the cgMLST data with GrapeTree [3] using MSTreeV2. Number of allele differences are denoted as numbers on branches between nodes. The branches have been collapsed on ten differences or less. The number inside each node is the sequence type based on the Warwick 7-gene MLST scheme [4].
```{r, fig.width=15, fig.height=12}
include_graphics("../figures/GrapeTree_cgMLST.png")
```

@@

broiler_cgMLST <- read.table("../data/chewbbaca/complete_results/cgMLST_broiler.tsv",
                           sep = "\t",
                           header = TRUE,
                           colClasses = "factor",
                           row.names = 1)

pig_cgMLST <- read.table("../data/chewbbaca/complete_results/cgMLST_pig.tsv",
                           sep = "\t",
                           header = TRUE,
                           colClasses = "factor",
                           row.names = 1)

fox_cgMLST <- read.table("../data/chewbbaca/complete_results/cgMLST_fox.tsv",
                           sep = "\t",
                           header = TRUE,
                           colClasses = "factor",
                           row.names = 1)

bird_cgMLST <- read.table("../data/chewbbaca/complete_results/cgMLST_bird.tsv",
                           sep = "\t",
                           header = TRUE,
                           colClasses = "factor",
                           row.names = 1)

@@

ggplot(complete_data, aes(Percent)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.2, color = "black",
                fill = "grey80") +
  stat_function(fun = dnorm,
                args = with(complete_data, c(mean = mean(Percent), sd = sd(Percent))),
                color = "red") +
  geom_segment(aes(x = complete_data$Percent[1],
                   xend = complete_data$Percent[1], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#4575b4") +
  geom_segment(aes(x = complete_data$Percent[2],
                   xend = complete_data$Percent[2], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#74add1") +
  geom_segment(aes(x = complete_data$Percent[3],
                   xend = complete_data$Percent[3], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#f46d43") +
  geom_segment(aes(x = complete_data$Percent[4],
                   xend = complete_data$Percent[4], y = 0.08, yend = 0),
               arrow = arrow(length = unit(0.3, "cm"),
                             type = "closed"),
               color = "#fdae61") +
  geom_text(label = "p < 0.001",
            x = 83.30,
            y = 0.16,
            angle = 90) +
  geom_vline(xintercept = mean(complete_data$Percent),
             size = 1) +
  labs(x = "Percent (%) distances < 0.1 or > 0.8",
       y = "Density") +
  ggtitle(label = "Distribution of percentages",
          subtitle = "Number of iterations: 1000") +
  theme(plot.title = element_text(hjust = 0))

@@

# References
[1] Alikhan N-F et al. A genomic overview of the population structure of Salmonella. Casadesús J, editor. PLOS Genet [Internet]. 2018 Apr 5;14(4):e1007261. Available from: https://dx.plos.org/10.1371/journal.pgen.1007261

[2] Machado MP et al. chewBBACA: A complete suite for gene-by-gene schema creation and strain identification. Microb Genomics. 2018;1–7.

[3] Z Zhou et al. (2018) “GrapeTree: Visualization of core genomic relationships among 100,000 bacterial pathogens”, Genome Res. doi: https://doi.org/10.1101/gr.232397.117

[4] Wirth T et al. (2006) "Sex and virulence in Escherichia coli : an evolutionary perspective", Molecular Microbiology (2006) 60 (5), 1136–1151, doi: 10.1111/j.1365-2958.2006.05172.x
